The Ky Fan Norms and Beyond: Dual Norms and Combinations for Matrix Optimization

### 2025-11-28 article_final/main.tex

Our experiments reveal important insights into the role of matrix norms in optimization. First, using the example of Neon (the rank-one Fanion), we show that not every LMO-based algorithm is effective, despite sharing the same asymptotics in the general bounds of \citet{kovalev2025understanding} and \citet{riabinin2025gluon}. This suggests that existing theoretical guarantees should be refined to better explain empirical performance.

@@

\paragraph{Schatten norms}
\citet{cesista2025schattenp} considered Schatten-$p$ norms:
\[\norm{\mM^t}_{S_p} = \left(\sum_{i=1}^{\min(m, n)}\sigma_i^p\right)^{1/p}\,,
\]
which produce the update
\[
\mX^{t+1} = \mX^t - \gamma_t \mU \frac{\diag\left(\sigma_1^{q-1},\dots, \sigma_{\min(m,n)}^{q-1}\right)}{\left(\sum_{i=1}^{\min(m,n)}{\sigma_i^q}\right)^{\frac{q-1}{q}}}\mV^\top
\]
where $p$ and $q$ satisfy $\frac{1}{p} + \frac{1}{q} = 1$. This formula recovers Neon when $p\rightarrow1$ (provided $\sigma_1 > \sigma_2$, which holds on real data), NSGD when $p=2$, and Muon when $p\rightarrow\infty$.

@@

Deriving the update for arbitrary $k$: Since $\normkfk{\cdot}^\dagger = \max\{\frac{1}{k} \normn{\cdot}, \norms{\cdot}\}$ (see \citet{bhatia2013matrix}, p. 96), the goal is to reach $\max\{\frac{1}{k}\sum_{i=1}^{\min(m,n)}\sigma_i, \sigma_1\}$. The Neon update $\Delta = \vu_1 \vv_1^\top$ achieves $\sigma_1$, while the scaled Muon update $\Delta = \frac{1}{k}\sum_{i=1}^{\min(m,n)}\vu_i \vv_i^\top$ achieves $\frac{1}{k}\sum_{i=1}^{\min(m,n)}\sigma_i$. Thus, the update is either
\[
\mX^{t+1} = \mX^{t} - \gamma_t \vu_1 \vv_1^\top\text{ or }\mX^{t+1} = \mX^{t} - \frac{\gamma_t}{k}\mU \mV^\top\,,
\]
depending on which better minimizes the linear function $L(\mX) \equiv F(\mX^t) + \mM^t(\mX - \mX^t)$.

@@

\paragraph{Connection to existing algorithms}
The Fanion family unifies several known algorithms as special cases:
\begin{itemize}
	\item \textbf{Neon} is Fanion-1 (rank-one updates)
	\item \textbf{Muon} is Fanion-$\min\{m,n\}$ (full-rank updates)
	\item \textbf{Dion} (unsharded): The rank-$r$ Dion (Algorithm 1 from \citet{ahn2025dion}) without error feedback and without scaling of the update is actually Fanion-$r$ (see \citet{pethick2025understandingdion}, where Dion is written in a notation more similar to ours)
\end{itemize}

@@

\subsection{General Case}
First, we need a well-known fact mentioned, for example, in \citet[Table 1]{yu2012arithmetic}. For the sake of completeness, we provide a proof here.

@@

First, we compare F-Fanions on the following convex $L$-smooth problem:
\begin{equation}\label{eq:lls}
  F(\mX) = \frac{1}{2} \<(\mX-\mS), \mM (\mX - \mS) \mN> \to \min_{\mX \in \Rmn}\,
\end{equation}
where $\mX \in \Rmn$, $m=500$ and $n=500$ are typical dimensions of a neural network weight matrix, $\mS \in \Rmn$, $\mM \in \mathbb{S}^m_{+}$ and $\mN \in \mathbb{S}^n_{+}$ are positive-semidefinite matrices. The spectra of $\mM$ and $\mN$ are uniformly distributed in the $(0, 1)$ interval. We set $\mS=0$, and $\mX^0 \sim 0.1 \cN(0, \mI)$.

@@

One cannot but notice the similarity between the updates of LMO optimizers in the vector $l_p$ and the matrix Schatten $S_p$ norms, which is illustrated by \tableref{tbl:mat_vs_vec_lmo}. The parallels are not limited to mere similarity of the updates and can be traced even in the empirical observations. SignSGD is very close to Adam, as noted in \citet{bernstein2024oldoptimizernewnorm}, and both Adam and Muon perform well in training large models \citep{zhao2025deconstructing, liu2025muon}. NSGD is the same for both matrix and vector cases. Greedy Coordinate Descent is not applied to high-dimensional problems, so from this perspective it is not surprising that one-rank Neon underperforms on such problems.

@@

\paragraph{Improvements of Muon} Regarding the first point, a large number of applications and improvements of Muon has been proposed in less than a year. \citet{liu2025muon} adapted the algorithm for training language models larger than NanoGPT. \citet{shah2025practical} organized efficient hyperparameter transfer by combining Muon with maximal update parametrization. To construct their COSMOS optimizer, \citet{chen2025cosmoshybridadaptive} applied computationally intensive updates of SOAP optimizer \citep{vyas2025soap} to a low-dimensional ``leading eigensubspace'' while using memory-efficient methods like Muon for the remaining parameters. \citet{amsel2025polar, grishina2025chebyshev} proposed more efficient alternatives to Newton-Schulz algorithm. \citet{si2025adamuon} introduced AdaMuon which combines element-wise adaptivity with orthogonal updates. We suppose that the described above or similar techniques can be applied to our optimizers as well. For example, F-Muon and S-Muon also benefit from faster alternatives to Newton-Schulz iterations, and Fanions may be a great substitute to Muon in COSMOS, because, as we have shown in \sectionref{sec:matrix-side}, Lanczos algorithm is much faster than Newton-Schulz iterations on very large matrices.

\paragraph{Theory behind Muon} Regarding the second point, there has been a prolonged gap in theory behind Muon, simplistic derivation of \citet{bernstein2025deriving} based on \citet{bernstein2024oldoptimizernewnorm} excluded. This gap, as it seems to us, is not even now completely closed. For example, \citet{kovalev2025understanding} has provided convergence guarantees of Muon and in various settings, from which, however, Muon's supremacy cannot be recovered. Indeed, although the obtained bounds depend on the norm choice, the asymptotics of the convergence remain the same as for NSGD and other optimizers, $K = \cO(\epsilon^{-4})$ in an $L$-smooth stochastic case.

A similar drawback has the article \citet{riabinin2025gluon}, where $L$-smoothness assumption is replaced with a more practical $(L_0, L_1)$-smoothness. By estimating smoothness and substituting it into their Theorem 1, the authors recovered the optimal fine-tuned stepsizes reported by \citet{pethick2025training}. However, they have not showed the optimality of the spectral or the RMS-to-RMS norm, which is observed in practice, as our comparison with NSGD highlights.

@@

\paragraph{The LMO and Error Feedback} We already mentioned that rank-$k$ unsharded Dion without error feedback and scaling of the update is Fanion-$k$. Since error feedback for Dion is very important as discovered by the ablation study in \citet{ahn2025dion}, F-Fanions and S-Fanions will benefit from it as well. In the context of federated learning, error feedback is effective even for compressed Muon \citep{gruntkowska2025efmuon}. The advantage of Fanions and S-Fanions is that less bits are required for transmission: $\sum_{i=1}^k \vu_i \vv_i^\top$ is easily transmitted as $\{\vu_1, \vv_1, \dots, \vu_k, \vv_k\}$ ($(m+n)\times k$ floats), while the sign part is as usual coded in $m \times n$ bits. Thus, the compression is ``built-in''. Moreover, there is an intriguing possibility to construct differentially private Fanions and S-Fanions with the specific more optimal non-Gaussian noise, as was done with DP-signSGD \citep{jang2024dplogsign}. This we leave for future research.

\paragraph{The nuclear norm in the LMO} As we found out only when transforming our results into an article, the nuclear norm has already been explored in the context of the linear minimization oracle. \citet{pethick2025sam} applied it to create $\nu$SAM, a new sharpness-aware minimization technique. It would be interesting to substitute $\normn{\cdot}$ with $\normkfk{\cdot}^\dagger$ in their approach. Since the SAM-neighborhood becomes more diverse, $k > 1$ might add to the accuracy boost, while preserving a small memory footprint and a small time overhead, if Dion-style power iterations are used.

\paragraph{Orthogonal research} Fanions, F-Fanions, and S-Fanions benefit from the general theoretical description of \citet{riabinin2025gluon}, and better learning rates can be predicted by calculating the trajectory smoothness. They could be transformed to Drop-Fanions by updating only some layers as in \citet{gruntkowska2025dropmuon}. They could be viewed as approximations of the Non-Euclidean Broximal Point Method for the corresponding norms \citep{gruntkowska2025noneuclideanbroximal}. One can clip them to produce ClippedScion-like algorithms \citep{pethick2025generalizedgradientnormclipping}. They could be made more memory-effective by the zero-order techniques that worked for Muon \citep{petrov2025zeromuon}. Finally, the results from \citet{shulgin2025inexactmuon} can be used to explain the robustness of Muon to the norm change that occurred in our experiments and to theoretically derive faster yet working approximate schemes to calculate the LMO; power iterations with a very limited number of iterations is a good algorithm to consider for the analysis.

@@

IO suggested using the nuclear norm in the \citet{bernstein2024oldoptimizernewnorm} framework. DM presented the problem at the MIPT optimization course. IK suggested using composite norms (though not the ones that induce F-Fanions and S-Fanions) and refactored \sectionref{sec:conic_combo}. NK suggested Lanczos algorithm as the most precise means to compute Fanions' updates, conducted experiments to prove this, and wrote \sectionref{sec:matrix-side}. AV conducted the finetuning of NanoGPT on Tiny Stories and wrote \sectionref{subsec:finetune}. All other work was done by AK: constructing Fanions, F-Fanions, S-Fanions, experimenting on the Linear Least Squares, CIFAR-airbench and NanoGPT pretrain, and writing the manuscript.

@@

In fact, those balls can be described easier if we use the results from \citet{yu2012arithmetic}. The Ky Fan $2$-norm ball is an intersection of three $l_1$ balls in $(x,y)$, $(x, z)$, and $(y,z)$ spaces. The 1-ball in the dual Ky Fan $2$-norm is an intersection of 1-ball the in $l_\infty$ norm and $\frac{1}{2}$-ball in the $l_1$ norm.

@@

\paragraph{NanoGPT pretrain} We used the standard setting of 8 $\times$ H100 as documented in \citet{modded_nanogpt_2024}.

>>>>

Our experiments reveal important insights into the role of matrix norms in optimization. First, using the example of Neon (the rank-one Fanion), we show that not every LMO-based algorithm is effective, despite sharing the same asymptotics in the general bounds of \citep{kovalev2025understanding} and \citep{riabinin2025gluon}. This suggests that existing theoretical guarantees should be refined to better explain empirical performance.

@@

\paragraph{Schatten norms}
\citep{cesista2025schattenp} considered Schatten-$p$ norms:
\[\norm{\mM^t}_{S_p} = \left(\sum_{i=1}^{\min(m, n)}\sigma_i^p\right)^{1/p}\,,
\]
which produce the update
\[
\mX^{t+1} = \mX^t - \gamma_t \mU \frac{\diag\left(\sigma_1^{q-1},\dots, \sigma_{\min(m,n)}^{q-1}\right)}{\left(\sum_{i=1}^{\min(m,n)}{\sigma_i^q}\right)^{\frac{q-1}{q}}}\mV^\top
\]
where $p$ and $q$ satisfy $\frac{1}{p} + \frac{1}{q} = 1$. This formula recovers Neon when $p\rightarrow1$ (provided $\sigma_1 > \sigma_2$, which holds on real data), NSGD when $p=2$, and Muon when $p\rightarrow\infty$.

@@

Deriving the update for arbitrary $k$: Since $\normkfk{\cdot}^\dagger = \max\{\frac{1}{k} \normn{\cdot}, \norms{\cdot}\}$ (see \citep{bhatia2013matrix}, p. 96), the goal is to reach $\max\{\frac{1}{k}\sum_{i=1}^{\min(m,n)}\sigma_i, \sigma_1\}$. The Neon update $\Delta = \vu_1 \vv_1^\top$ achieves $\sigma_1$, while the scaled Muon update $\Delta = \frac{1}{k}\sum_{i=1}^{\min(m,n)}\vu_i \vv_i^\top$ achieves $\frac{1}{k}\sum_{i=1}^{\min(m,n)}\sigma_i$. Thus, the update is either
\[
\mX^{t+1} = \mX^{t} - \gamma_t \vu_1 \vv_1^\top\text{ or }\mX^{t+1} = \mX^{t} - \frac{\gamma_t}{k}\mU \mV^\top\,,
\]
depending on which better minimizes the linear function $L(\mX) \equiv F(\mX^t) + \mM^t(\mX - \mX^t)$.

@@

\paragraph{Connection to existing algorithms}
The Fanion family unifies several known algorithms as special cases:
\begin{itemize}
	\item \textbf{Neon} is Fanion-1 (rank-one updates)
	\item \textbf{Muon} is Fanion-$\min\{m,n\}$ (full-rank updates)
	\item \textbf{Dion} (unsharded): The rank-$r$ Dion (Algorithm 1 from \citep{ahn2025dion}) without error feedback and without scaling of the update is actually Fanion-$r$ (see \citep{pethick2025understandingdion}, where Dion is written in a notation more similar to ours)
\end{itemize}

@@

\subsection{General Case}
First, we need a well-known fact mentioned, for example, in \citep[Table 1]{yu2012arithmetic}. For the sake of completeness, we provide a proof here.

@@

First, we compare F-Fanions on the following convex $L$-smooth problem:
\begin{equation}\label{eq:lls}
  F(\mX) = \frac{1}{2} \<(\mX-\mS), \mM (\mX - \mS) \mN> \to \min_{\mX \in \Rmn}\,
\end{equation}
where $\mX \in \Rmn$, $m=500$ and $n=500$ are typical dimensions of a neural network weight matrix, $\mS \in \Rmn$, $\mM \in \mathbb{S}^m_{+}$ and $\mN \in \mathbb{S}^n_{+}$ are positive-semidefinite matrices. The spectra of $\mM$ and $\mN$ are uniformly distributed in the $(0, 1)$ interval. We set $\mS=0$, and elements of $\mX^0$ are instances of random value with distribution $\sim 0.1 \cN(0, 1)$.

@@

One cannot but notice the similarity between the updates of LMO optimizers in the vector $l_p$ and the matrix Schatten $S_p$ norms, which is illustrated by \tableref{tbl:mat_vs_vec_lmo}. The parallels are not limited to mere similarity of the updates and can be traced even in the empirical observations. SignSGD is very close to Adam, as noted in \citep{bernstein2024oldoptimizernewnorm}, and both Adam and Muon perform well in training large models \citep{zhao2025deconstructing, liu2025muon}. NSGD is the same for both matrix and vector cases. Greedy Coordinate Descent is not applied to high-dimensional problems, so from this perspective it is not surprising that one-rank Neon underperforms on such problems.

@@

\paragraph{Improvements of Muon} Regarding the first point, a large number of applications and improvements of Muon has been proposed in less than a year. \citep{liu2025muon} adapted the algorithm for training language models larger than NanoGPT. \citep{shah2025practical} organized efficient hyperparameter transfer by combining Muon with maximal update parametrization. To construct their COSMOS optimizer, \citep{chen2025cosmoshybridadaptive} applied computationally intensive updates of SOAP optimizer \citep{vyas2025soap} to a low-dimensional ``leading eigensubspace'' while using memory-efficient methods like Muon for the remaining parameters. \citep{amsel2025polar, grishina2025chebyshev} proposed more efficient alternatives to Newton-Schulz algorithm. \citep{si2025adamuon} introduced AdaMuon which combines element-wise adaptivity with orthogonal updates. We suppose that the described above or similar techniques can be applied to our optimizers as well. For example, F-Muon and S-Muon also benefit from faster alternatives to Newton-Schulz iterations, and Fanions may be a great substitute to Muon in COSMOS, because, as we have shown in \sectionref{sec:matrix-side}, Lanczos algorithm is much faster than Newton-Schulz iterations on very large matrices.

\paragraph{Theory behind Muon} Regarding the second point, there has been a prolonged gap in theory behind Muon, simplistic derivation of \citep{bernstein2025deriving} based on \citep{bernstein2024oldoptimizernewnorm} excluded. This gap, as it seems to us, is not even now completely closed. For example, \citep{kovalev2025understanding} has provided convergence guarantees of Muon and in various settings, from which, however, Muon's supremacy cannot be recovered. Indeed, although the obtained bounds depend on the norm choice, the asymptotics of the convergence remain the same as for NSGD and other optimizers, $K = \cO(\epsilon^{-4})$ in an $L$-smooth stochastic case.

A similar drawback has the article \citep{riabinin2025gluon}, where $L$-smoothness assumption is replaced with a more practical $(L_0, L_1)$-smoothness. By estimating smoothness and substituting it into their Theorem 1, the authors recovered the optimal fine-tuned stepsizes reported by \citep{pethick2025training}. However, they have not showed the optimality of the spectral or the RMS-to-RMS norm, which is observed in practice, as our comparison with NSGD highlights.

@@

\paragraph{The LMO and Error Feedback} We already mentioned that rank-$k$ unsharded Dion without error feedback and scaling of the update is Fanion-$k$. Since error feedback for Dion is very important as discovered by the ablation study in \citep{ahn2025dion}, F-Fanions and S-Fanions will benefit from it as well. In the context of federated learning, error feedback is effective even for compressed Muon \citep{gruntkowska2025efmuon}. The advantage of Fanions and S-Fanions is that less bits are required for transmission: $\sum_{i=1}^k \vu_i \vv_i^\top$ is easily transmitted as $\{\vu_1, \vv_1, \dots, \vu_k, \vv_k\}$ ($(m+n)\times k$ floats), while the sign part is as usual coded in $m \times n$ bits. Thus, the compression is ``built-in''. Moreover, there is an intriguing possibility to construct differentially private Fanions and S-Fanions with the specific more optimal non-Gaussian noise, as was done with DP-signSGD \citep{jang2024dplogsign}. This we leave for future research.

\paragraph{The nuclear norm in the LMO} As we found out only when transforming our results into an article, the nuclear norm has already been explored in the context of the linear minimization oracle. \citep{pethick2025sam} applied it to create $\nu$SAM, a new sharpness-aware minimization technique. It would be interesting to substitute $\normn{\cdot}$ with $\normkfk{\cdot}^\dagger$ in their approach. Since the SAM-neighborhood becomes more diverse, $k > 1$ might add to the accuracy boost, while preserving a small memory footprint and a small time overhead, if Dion-style power iterations are used.

\paragraph{Orthogonal research} Fanions, F-Fanions, and S-Fanions benefit from the general theoretical description of \citep{riabinin2025gluon}, and better learning rates can be predicted by calculating the trajectory smoothness. They could be transformed to Drop-Fanions by updating only some layers as in \citep{gruntkowska2025dropmuon}. They could be viewed as approximations of the Non-Euclidean Broximal Point Method for the corresponding norms \citep{gruntkowska2025noneuclideanbroximal}. One can clip them to produce ClippedScion-like algorithms \citep{pethick2025generalizedgradientnormclipping}. They could be made more memory-effective by the zero-order techniques that worked for Muon \citep{petrov2025zeromuon}. Finally, the results from \citep{shulgin2025inexactmuon} can be used to explain the robustness of Muon to the norm change that occurred in our experiments and to theoretically derive faster yet working approximate schemes to calculate the LMO; power iterations with a very limited number of iterations is a good algorithm to consider for the analysis.

@@

IO suggested using the nuclear norm in the \citep{bernstein2024oldoptimizernewnorm} framework. DM presented the problem at the MIPT optimization course. IK suggested using composite norms (though not the ones that induce F-Fanions and S-Fanions) and refactored \sectionref{sec:conic_combo}. NK suggested Lanczos algorithm as the most precise means to compute Fanions' updates, conducted experiments to prove this, and wrote \sectionref{sec:matrix-side}. AV conducted the finetuning of NanoGPT on Tiny Stories and wrote \sectionref{subsec:finetune}. All other work was done by AK: constructing Fanions, F-Fanions, S-Fanions, experimenting on the Linear Least Squares, CIFAR-airbench and NanoGPT pretrain, and writing the manuscript.

@@

In fact, those balls can be described easier if we use the results from \citep{yu2012arithmetic}. The Ky Fan $2$-norm ball is an intersection of three $l_1$ balls in $(x,y)$, $(x, z)$, and $(y,z)$ spaces. The 1-ball in the dual Ky Fan $2$-norm is an intersection of 1-ball the in $l_\infty$ norm and $\frac{1}{2}$-ball in the $l_1$ norm.

@@

\paragraph{NanoGPT pretrain} We used the standard setting of 8 $\times$ H100 as documented in \citep{modded_nanogpt_2024}.
